Nuclear Wastes: we're kinda doomed

### 2017-01-23 paper.tex
spell checked

\section{Introduction}
Nuclear power continues to play an important role in global energy supply.
The International Atomic Energy Agency (IAEA) estimated that the power generated by nuclear power
is continuing to grow and will increase by $40\%$ by 2030.\cite{iaea12}
Whereas the ``front end'' of the nuclear fuel cycle, for example reactor physics and power output, draws
the most attention from the public, the ``back end'' of the fuel cycle, for example waste management and
disposal, deserves equal, if not more, attention.
Currently, the IAEA estimates that there are over 10,000 metric tons of highly radioactive nuclear waste 
generated worldwide annually,
most of which is spent nuclear fuel from nuclear reactors.\cite{iaea08, r12}
The world will have over $373,800$ metric tons of highly radioactive waste stock by the end of 2020.\cite{r12}
Highly radioactive waste remains hazardous for up to $10^6$ years after it is produced and thus requires extremely
careful 
Nevertheless, there is currently no permanent repository in the world to dispose of these highly radioactive
waste.~\footnote{The Waste Isolation Pilot Plant (WIPP) in New Mexico is designed to be a permanent radioactive
    waste repository, but only
for waste classified as Transuranic, which, by the Nuclear Regulatory Commission standard, does not have particularly
high radioactivity but only long half-life (See Sec.~\ref{sec:tru}). In particular, WIPP does not accept spent nuclear
fuel.}  More than $75\%$ of the spent nuclear fuel worldwide is stored onsite at the nuclear power plants in
water pools that are close to their limiting capacity.\cite{aa12}

Two weeks after the davastating earthquake and tsunami hit the Fukushima Daiichi Nuclear Power Plant in Japan,
the water pool cooling system was cut off from power and officials feared that
the highly radioactive waste in the pools would catch fire and send radioactive smoke across a much larger part of Japan, including Tokyo.
The Nuclear Regulatory Comission (NRC), the independent government agency of the U.S. government overseeing 
nuclear energy safety,
estimates that a major fire in a U.S. spent fuel pool is expected to
displace 3.4 million people in an area larger than New Jersy.\cite{s16}

As we can see, nuclear waste management and disposal is becoming a increasingly urgent matter globally.
In light of the high stake,
the uncertainty of the stability of a geological repository over a extremely long timescale has evolved into 
a topic not only of scientific contention but also of policy debate.
How well is the predicative power of science? How much trust should we put into the uncertainty estimates
(i.e. how well can science predict its predicative power)?
In this work, we will attempt to examine some of the scientific issues underlying nuclear waste management
and disposal. We will start by building up some
language and knowledge necessary to understand how nuclear waste is generated (Section~\ref{sec:phys}),
then apply those principles to study
what physical properties nuclear waste possesses (Section~\ref{sec:waste}), its management and temporary storage
(Section~\ref{sec:temp}), and its disposal (Section~\ref{sec:disposal}).

\section{Radioactivity and nuclear fission}
\label{sec:phys}
    Here we detail the nuclear physics underlying most of the things we will talk about
    in this work. We will apply these results in our discussion of nuclear wastes.
    \subsection{Radioactivity}
    \subsubsection{Nuclear shell model vs liquid drop model(?)}
        By solving Schr\"{o}dinger's equation\cite{k88}
    \subsubsection{Binding Energy(?)}

        \subsubsection{Quantum theory of radioactive decays}
        We have learned that radioactivity is stochastic in nature and radioactive decay follows
        an exponential law
        \begin{equation} \label{eq:exp}
            \lambda = -\frac{(dN/dt)}{N},
        \end{equation}
        where $\lambda$ is the decay constant, and the right hand side the probability of decay
        per time for an atom. The fundamental assumption is that the probability of decay is constant,
        regardless of the age of the atom. This ``memorylessness'' assumption naturally leads to the
        exponential law of radioactive decay
        \begin{equation}
            N(t) = N_0 e^{-\lambda t},
        \end{equation}
        where $N_0$ gives the number of nuclei at $t=0$.

        Define activity $\mathcal{A}(t)$ to be the rate of decays in a sample
        \begin{equation} \label{eq:act}
            \mathcal{A}(t)\equiv \lambda N(t).
        \end{equation}
        Here we outline an quantum mechanics argument for an exponential decay law that is perhaps more
        convincing.
        We approximate the potential by $V+V_p$ where $V$ is the nuclear potential and $V_p$ is a small
        time-dependent perturbation that causes the transition between the states. We know that there has
        to be time dependence in the potential, otherwise for a superposition of eigenstates
        $\ket{psi}=c_1\ket{\psi_1} + c_2\ket{\psi_2}$, the coefficients $c_1$ and $c_2$ would 
        have to be constant and
        persistent decay of one state into the other would not happen.
        Fermi's Golden Rule \#2 (which is derived in appendix \ref{a:fermi})
        gives the transition rate
        \begin{equation}
            \label{eq:trans}
            \lambda = \frac{2\pi}{2\hbar}|\bra{\psi_f}V_p\ket{\psi_i}|^2\rho(E_f),
        \end{equation}
        where $\ket{\psi_i}$ denotes the initial state , $\bra{\psi_f}$ the final state, $V_p$ the perturbation
        potential and $\rho(E_f)$ the density of state of the final state, defined by
        \begin{equation}
            \rho(E_f) = \frac{dn_f}{dE_f}.
        \end{equation}
        Note that as in perturbation theory, $\ket{\psi_i}$ and
        $\ket{\psi_f}$ are both eigenstates of the nuclear potential $V$. Therefore, as long as
        the decay probability is small, the result from perturbation theory should hold. Since
        there is no time dependency in equation~\ref{eq:trans},the
        decay rate should be independent of time, which is the fundamental assumption made in
        equation~\ref{eq:exp}.

        \subsubsection{Quantifying radioactivity}
        The radioactivity of a sample $\mathcal{A}$ (defined in equation~\ref{eq:act}) 
        is the number of decays of the sample per unit time. The SI
        unit for radioactivity is becquerel (Bq) and it equals one decay per second. Another commonly used
        unit is curie (Ci, with $1\text{ Ci}=3.7\times 10^{10}\text{ Bq}$) and it is defined as
        the activity of radium-266. A fission product, cesium-137 has an activity of $3.215\times 10^{12}$ Bq
        per gram. By contrast, the radioactivity of a human body (mainly due to the small amount of 
        potassium-40 and carbon-14) is about $8000$ Bq.

@@

        \subsubsection{Fission and fission products}
        Conjectured by Bohr and proven by Alfred O. Nier in 1939,
        the fission of uranium-235 induced by a thermal neutron
        is the best-known example of induced fission. 
        The neutron capture produces a compound nucleus ${}^{236}U$ that is in an excited state.
        Sometimes the nucleus can decay by $\gamma$ emission in capture reaction. \red{(?)}
        But most of the time, the excitation energy deforms the nucleus (see FIG. \ref{fig:fission}).
        When the
        deformation reaches a certain point, the Coulomb force overcomes the short-ranged nuclear
        force (residual strong force) and the nucleus disintegrate into two large fragments and several neutrons. The
        positively-charged fragments(fission products) attain approximately $170$ MeV of kinetic energy as
        Coulomb force continues to drive them apart.\cite{l01}
        A typical reaction is described by
        \[{}^{235}_{92}\text{U} + n \rightarrow {}^{92}_{36}\text{Kr} + {}^{142}_{56}\text{Ba} + 2 n.\]
        The energy liberated is about $180$ MeV. The combination of fission products is not
        unique. \cite{w98, gc01}
        Very often, photons, electrons and neutrinos are also emitted through beta and gamma decay.
        \red{(insert 10.2 from Lilley here? distribution of fission fragment masses)}
        Radioactive if neutron-to-proton ratio is over 5. On average, $2.5$ neutrons are produced per fission.\cite{l01} 
        \begin{figure}[h]
            \centering
            \includegraphics[width=0.8\textwidth]{fission.png}
            \caption{Illustration of the fission process for uranium-235.\cite{l01}}
            \label{fig:fission}
        \end{figure}

@@

\section{Nuclear waste: classification and properties}
\label{sec:waste}
    \red{how much of each is generated and what the impact is; industry vs military}\\
    Different organizations have slightly different classification schemes for nuclear waste.
    However, their definitions of high-level waste (HLW), the most detrimental kind, are relatively consistent.
    The International Atomic Energy Agency (IAEA) classifies nuclear waste by considering the activity
    content and the half-life of the radionuclides (see FIG. \ref{fig:scheme}).
    Activity content is a measure of how much radioactivity
    a given amount of waste contains: a form of waste may possesses a high activity content due to
    \red{a high concentration of radionuclides or the presence of radionuclides with high activity.}\cite{iaea09}
    Each form of waste, depending on the half-life, activity content, and temperature, requires different
    disposal methods.
    \begin{figure}[h]
        \centering
        \includegraphics[width=0.7\textwidth]{wastescheme.png}
        \caption{An illustration of the IAEA classification scheme for nuclear waste along with suggested
        disposal methods.\cite{iaea09}}
        \label{fig:scheme}
    \end{figure}
    We will focus on high-level waste (HLW, defined below) in the rest of this work.
    In this section, we will first examine the composition and properties of HLW and then
    comment on other categories of waste and disposal methods.

@@

    \section{Spent fuel management}
    \label{sec:temp}
    Spent fuel from nuclear reactors constitutes the major source of HLW. Spen fuel pools at the
    reactors hold spen fuel cores to cool them down and isolate them for a few years. After that,
    reprocessing can occur to extract the most radioactive plutonium and uranium isotopes and
    convert the waste into more managable form (vitrification). However, reprocessing of commercial
    nuclear spent fueld did not
    happen in this country for various historical reasons. 
    \subsection{Spent fuel pools}
    \red{boron-10 and radiation shielding.}
    The heat output of the spent fuel decreases by $99\%$ after the first year and by another factor of $5$
    by the 5-year mark.\cite{aa12} Therefore, short-term cooling of spent fuel is crucial before
    futher disposal or reprocessing can be done.
    US nuclear power plants generally have an operation cycle of about 2 years. At the end of each operation cycle,
    one-quater to one-third of its fuel assemblies (spent fuel) are removed.
    Most of the nuclear industry's spent fuel is now stored in spent fuel water pools ``at-reactor (AR)'',
    which means either within the reactor building or in an ajacent spent fuel building linked to the reactor
    AR pools minimize the difficulty of transporting spent fuel.\cite{iaea99}
    US has generated about $65,000$ metric tons of spent fuel, of which $75\%$ is stored in AR pools.\cite{a11}
    by a transfer tunnel. 
    Spent fuel pools  are usually about 15-meter deep borated water pools, with storage rack at the bottom to hold the fuel assemblies.
    The water absorbs the heat and shields the radiation. Water is the chosen medium because it is cheap and
    can provide coooling by circulation. Moreover, water can serve as a shielding from radiation, because it has a
    fairly high density and hence forces the radiated particles (especially the neutrons) to interact with
    the water molecules and lose energy \red{(more precise terms needed, cladding materials)}. Water temperature is maintained
    below 50 $^\circ$C via circulation and cooling. The integrity of the storage can persist well beyond 
    50 years.\cite{a11, iaea99}

@@

    Almost all of the reprocessing plants use the PUREX (Plutonium Uranium Redox EXtraction) process
    to reprocess spent fuel. This process utilizes the differences in oxidation potential 
    between the heavy elements (uranium and plutonium) from the other lighter fission products.
    The process involves three steps: (a) adding appropriate solvent (usually nitric acid) to absorb and oxidize 
    plutonium and uranium while leaving out other elements;
    (b) adding a reducing (de-oxidizing) agent that reduces (de-oxidizes) only the
    oxidation state of plutonium to one that is insoluble
    and thus seperable; and finally (c) extracting the plutonium from the mixture. In this way
    the oxidized plutonium and uranium are separated and can be reduced back to reusable radionuclides.\cite{lb01} 
    The extracted plutonium then returns to the reactor and transforms into short-lived fission products
    by fission. Un-reprocessed fuel usually takes 10 times as much time to reach the safe
    activity level (\sim$10^5$ years) as reprocessed fuel (\sim$10^4$ years). Moreover, since fission products
    make up a small portion of the spent fuel, reprocessing greatly reduces the mass of the waste. The reprocessed
    waste can be turned into liquid form, mixed with frit (the substance from which glass is made) and then
    vitrified (made into glass). The vitrification process immobilizes the radioactive particles and makes
    the waste much more safe and manageable, but vitrification can only be applied to reprocessed spent fuel.

    It is worthwhile to note that the PUREX processed was designed specifically to seperate plutonium for
    bombs. Therefore, literally any other reprocessing approach would be less prone to proliferation.
    Researchers are actively investigating alternatives to the PUREX technology. For example, the UREX+ (for
    uranium extraction) technique extracts the uranium but leaves the plutonium mixed with other
    radioactive materials. Another popular alternative, pyrochemical reprocessing, inserts giant electrodes
    into the chemical bath of nuclear waste and electroplates the plutonium and other transuranics onto
    the electrode. However, none of these alternatives have been put into practice due to the high construction
    and operating costs of reprocessing facilities.\cite{aa12}

@@

    \subsection{Dry cask storage}
    Spent fuel pools have very limited capacities. 
    Overground dry cask storage provides a low-cost temporary storage before the spent fuel is disposed of.
    After cooling in the spent fuel pools for at least a year, the spent fuel can be stored in dry casks filled with inert gas. 
    The dry casks are often metal cyliders welded to prevent leakage. Each cask is surrounded by
    additional steel and concrete to provide shielding.
    The heat from the cask is removed by passive
    cooling (air circulation). The Nuclear Regulatory Comission is hoping that the casks will
    last safely for as long as 100 years, yet corrosions and cracking occur in 30 years
    or less.\cite{aa12} As such, without the prospect of a permanent geological repository,
    overground dry cask storage does not seem to be a long-term solution. The doubts were expressed
    dudring  the Nuclear Waste Technical Review Board 2009 meeting \cite{nwtb09, aa12}
    \begin{displayquote}
    Leaving the spent fuel onsite for extended periods of time was never intended and is not
    responsible. (Dry cask manufacturer)

    It is not ethical, basically to plan for long-term storage without pursuing a well defined repository
    program. (Department of Energy employee speaking as a citizen)
    \end{displayquote}
    Even more striking is the fact that more than $75\%$ of the spent fuel are still stored in densely
    packed pools
    that are close to or over their designed limits.\cite{a11,aa12}
    Therefore, there is clearly a need for a long-term disposal plan, which we will epxplore in
    the next section.\\
    \red{(A lot more drama with the NRC going on here, loss of credibility.)}

\section{Disposal of HLW}
\label{sec:disposal}
    The timeframe of nuclear waste disposal is about $10^4-10^6$ years, after which the radioactivity
    of the waste is predicted to decline to a safe level.
    Deep geological repositories that are stable over timescales naturally offer a good option for disposal.
    It is accepted that geological repositories in a stable environment over 300 meters below ground
    is safter than indefinite overground storage of HLW, reprocessed or not.\cite{fmr11}
    In this section, we first state the requirements for a HLW geological disposal solution, and then
    we look at  deep geological repository, and its alternatives (or variations): deep boreholes storage
    and reversible repository.
    \subsection{Requirements for HLW disposal}
    \subsection{Survey of possibilities}
    transmutation, ocean floor, deep geological disposal
    \subsection{Deep borehole disposal}
        \subsection{Deep geological repository: Yucca Mountain}
        \subsubsection{Risk}
    \subsection{Reversible disposal sites}

\section{Discussions}

k
\begin{acknowledgments}
Thanks Joel and peeps. Arjendu and M. V. Ramana.
\end{acknowledgments}

>>>>

\section{Introduction}
Nuclear power continues to play an important role in global energy supply.
The International Atomic Energy Agency (IAEA) estimated that the power generated by nuclear power
is continuing to grow and will increase by $40\%$ by 2030.\cite{iaea12}
Whereas the ``front end'' of the nuclear fuel cycle, such as reactor physics and power output, draws
the most attention from the public, the ``back end'' of the fuel cycle, like waste management and
disposal, deserves equal, if not more, attention.
Currently, the IAEA estimates that there are over 10,000 metric tons of highly radioactive nuclear waste 
generated worldwide annually,
most of which is spent nuclear fuel from nuclear reactors.\cite{iaea08, r12}
The world will have over $457,000$ metric tons of highly radioactive waste stock by the end of 2020.\cite{r12}
Highly radioactive waste remains hazardous for up to $10^6$ years after it is produced and thus
requires extremely careful disposal methods.
Nevertheless, there is currently no permanent repository in the world to dispose of these highly radioactive
waste.~\footnote{The Waste Isolation Pilot Plant (WIPP) in New Mexico is designed to be a permanent radioactive
    waste repository, but only
for waste classified as Transuranic, which, by the Nuclear Regulatory Commission standard, does not have particularly
high radioactivity but only long half-life (See Sec.~\ref{sec:tru}). In particular, WIPP does not accept spent nuclear
fuel.}

It is alarming that more than $75\%$ of the spent nuclear fuel worldwide is stored on-site at the nuclear power plants
in water pools that have exceeded their designed capacity and are reaching their limiting capacity.\cite{aa12}
Two weeks after the devastating earthquake and tsunami hit the Fukushima Daiichi Nuclear Power Plant in Japan,
the water pool cooling system was cut off from power. Officials feared that
the highly radioactive waste in the pools would catch fire and send radioactive smoke
across a much larger part of Japan, including Tokyo. For some ``fortuitous'' reasons, the fear did not come true.
However, the Nuclear Regulatory Commission (NRC), the independent government agency
of the U.S. government overseeing nuclear energy safety,
estimates that a major fire in a U.S. spent fuel pool is expected to
displace 3.4 million people in an area larger than New Jersey.\cite{s16}

As we can see, nuclear waste management and disposal is becoming a increasingly urgent matter globally.
In light of the high stake,
the uncertainty of the stability of a geological repository over a extremely long timescale has evolved into 
a topic not only of scientific contention but also of policy debate.
How well is the predicative power of science? How much trust should we put into the uncertainty estimates
(i.e. how well can science predict its predicative power)?

In this work, we will attempt to examine some of the scientific issues underlying nuclear waste management
and disposal. We will start by building up some
language and knowledge necessary to understand how nuclear waste is generated (Section~\ref{sec:phys}),
then apply those principles to study
what physical properties nuclear waste possesses (Section~\ref{sec:waste}), its management and temporary storage
(Section~\ref{sec:temp}), and the debates around its eventual disposal (Section~\ref{sec:disposal}).

\section{Radioactivity and nuclear fission}
\label{sec:phys}
    Here we detail the nuclear physics underlying most of the things we will talk about
    in this work. We will apply these results in our discussion of nuclear wastes.
    \subsection{Radioactivity}
    %\subsubsection{Nuclear shell model vs liquid drop model(?)}
    %\subsubsection{Binding Energy(?)}
        \subsubsection{Quantum theory of radioactive decays}
        We have learned that radioactivity is stochastic in nature and radioactive decay follows
        an exponential law
        \begin{equation} \label{eq:exp}
            \lambda = -\frac{(dN/dt)}{N},
        \end{equation}
        where $\lambda$ is the decay constant, and the right hand side the probability of decay
        per time for an atom. The fundamental assumption is that the probability of decay is constant,
        regardless of the age of the atom. This ``memorylessness'' assumption naturally leads to the
        exponential law of radioactive decay
        \begin{equation}
            N(t) = N_0 e^{-\lambda t},
        \end{equation}
        where $N_0$ gives the number of nuclei at $t=0$.

        Define activity $\mathcal{A}(t)$ to be the rate of decays in a sample
        \begin{equation} \label{eq:act}
            \mathcal{A}(t)\equiv \lambda N(t).
        \end{equation}
        Here we outline an quantum mechanics argument for an exponential decay law that is perhaps more
        convincing.
        We approximate the potential by $V+V_p$ where $V$ is the nuclear potential and $V_p$ is a small
        time-dependent perturbation that causes the transition between the states. We know that there has
        to be time dependence in the potential, otherwise for a superposition of eigenstates
        $\ket{psi}=c_1\ket{\psi_1} + c_2\ket{\psi_2}$, the coefficients $c_1$ and $c_2$ would 
        have to be constant and
        persistent decay of one state into the other would not happen.
        Fermi's Golden Rule \#2 (which is derived in appendix \ref{a:fermi})
        gives the transition rate
        \begin{equation}
            \label{eq:trans}
            \lambda = \frac{2\pi}{2\hbar}|\bra{\psi_f}V_p\ket{\psi_i}|^2\rho(E_f),
        \end{equation}
        where $\ket{\psi_i}$ denotes the initial state , $\bra{\psi_f}$ the final state, $V_p$ the perturbation
        potential and $\rho(E_f)$ the density of state of the final state, defined by
        \begin{equation}
            \rho(E_f) = \frac{dn_f}{dE_f}.
        \end{equation}
        In simple terms, $\rho(E_f)dE$ gives the proportion of the particles with energy in the
        really small interval between
        $E_f$ and $(E_f+dE)$.
        Note that as in perturbation theory, $\ket{\psi_i}$ and
        $\ket{\psi_f}$ are both eigenstates of the nuclear potential $V$. Therefore, as long as
        the decay probability is small, the result from perturbation theory should hold. Since
        there is no time dependency in Eq.~\ref{eq:trans},the
        decay rate should be independent of time, which is the fundamental assumption made in
        Eq.~\ref{eq:exp}.

        \subsubsection{Quantifying radioactivity}
        The radioactivity of a sample $\mathcal{A}$ (defined in Eq.~\ref{eq:act}) 
        is the number of decays of the sample per unit time. The SI
        unit for radioactivity is becquerel (Bq) and it equals one decay per second. Another commonly used
        unit is curie (Ci, with $1\text{ Ci}=3.7\times 10^{10}\text{ Bq}$) and it is defined as
        the activity of radium-266. A fission product, cesium-137 has an activity of $3.215\times 10^{12}$ Bq
        per gram. By contrast, the radioactivity of a human body (mainly due to the small amount of 
        potassium-40 and carbon-14) is about $8000$ Bq.

@@

        \subsubsection{Fission and fission products}
        Conjectured by Bohr and proven by Alfred O. Nier in 1939,
        the fission of uranium-235 induced by a thermal neutron
        is the best-known example of induced fission. 
        The neutron capture produces a compound nucleus ${}^{236}U$ that is in an excited state.
        Sometimes the nucleus can decay by $\gamma$ emission in capture reaction. \red{(?)}
        But most of the time, the excitation energy deforms the nucleus (see Fig. \ref{fig:fission}).
        When the
        deformation reaches a certain point, the Coulomb force overcomes the short-ranged nuclear
        force (residual strong force) and the nucleus disintegrate into two large fragments and several neutrons. The
        positively-charged fragments(fission products) attain approximately $170$ MeV of kinetic energy as
        Coulomb force continues to drive them apart.\cite{l01}
        A typical reaction is described by
        \[{}^{235}_{92}\text{U} + n \rightarrow {}^{92}_{36}\text{Kr} + {}^{142}_{56}\text{Ba} + 2 n.\]
        The energy liberated is about $180$ MeV. The combination of fission products is not
        unique. \cite{w98, gc01}
        Very often, photons, electrons and neutrinos are also emitted through beta and gamma decay.
    \red{(insert 10.2 from Lilley here? distribution of fission fragment masses)}
        Radioactive if neutron-to-proton ratio is over 5. On average, $2.5$ neutrons are produced per fission.\cite{l01} 
        \begin{figure}[h]
            \centering
            \includegraphics[width=0.8\textwidth]{fission.png}
            \caption{Illustration of the fission process for uranium-235.\cite{l01}}
            \label{fig:fission}
        \end{figure}

@@

\section{Nuclear waste: classification and properties}
\label{sec:waste}
    \red{how much of each is generated and what the impact is; industry vs military}\\
    Different organizations have slightly different classification schemes for nuclear waste.
    However, their definitions of high-level waste (HLW), the most detrimental kind, are relatively consistent.
    The International Atomic Energy Agency (IAEA) classifies nuclear waste by considering the activity
    content and the half-life of the radionuclides (see Fig. \ref{fig:scheme}).
    Activity content is a measure of how much radioactivity
    a given amount of waste contains: a form of waste may possesses a high activity content due to
    \red{a high concentration of radionuclides or the presence of radionuclides with high activity.}\cite{iaea09}
    Each form of waste, depending on the half-life, activity content, and temperature, requires different
    disposal methods.
    \begin{figure}[h]
        \centering
        \includegraphics[width=0.7\textwidth]{wastescheme.png}
        \caption{An illustration of the IAEA classification scheme for nuclear waste along with suggested
        disposal methods.\cite{iaea09}}
        \label{fig:scheme}
    \end{figure}
    We will focus on high-level waste (HLW, defined below) in the rest of this work.
    In this section, we will first examine the composition and properties of HLW and then
    comment on other categories of waste and disposal methods.

@@

    \section{Spent fuel management}
    \label{sec:temp}
    Spent fuel from nuclear reactors constitutes the major source of HLW. Spent fuel pools at the
    reactors hold spent fuel cores to cool them down and isolate them for a few years. After that,
    reprocessing can occur to extract the most radioactive plutonium and uranium isotopes and
    convert the waste into more manageable form (vitrification). However, reprocessing of commercial
    nuclear spent fuel did not
    happen in this country for various historical reasons. 
    \subsection{Spent fuel pools}
    \red{boron-10 and radiation shielding.}
    The heat output of the spent fuel decreases by $99\%$ after the first year and by another factor of $5$
    by the 5-year mark.\cite{aa12} Therefore, short-term cooling of spent fuel is crucial before
    further disposal or reprocessing can be done.
    US nuclear power plants generally have an operation cycle of about 2 years. At the end of each operation cycle,
    one-quarter to one-third of its fuel assemblies (spent fuel) are removed.
    Most of the nuclear industry's spent fuel is now stored in spent fuel water pools ``at-reactor (AR)'',
    which means either within the reactor building or in an adjacent spent fuel building linked to the reactor
    AR pools minimize the difficulty of transporting spent fuel.\cite{iaea99}
    US has generated about $65,000$ metric tons of spent fuel, of which $75\%$ is stored in AR pools.\cite{a11}
    by a transfer tunnel. 
    Spent fuel pools  are usually about 15-meter deep borated water pools, with storage rack at the bottom to hold the fuel assemblies.
    The water absorbs the heat and shields the radiation. Water is the chosen medium because it is cheap and
    can provide cooling by circulation. Moreover, water can serve as a shielding from radiation, because it has a
    fairly high density and hence forces the radiated particles (especially the neutrons) to interact with
    the water molecules and lose energy \red{(more precise terms needed, cladding materials)}. Water temperature is maintained
    below 50 $^\circ$C via circulation and cooling. The integrity of the storage can persist well beyond 
    50 years.\cite{a11, iaea99}

@@

    Almost all of the reprocessing plants use the PUREX (Plutonium Uranium Redox EXtraction) process
    to reprocess spent fuel. This process utilizes the differences in oxidation potential 
    between the heavy elements (uranium and plutonium) from the other lighter fission products.
    The process involves three steps: (a) adding appropriate solvent (usually nitric acid) to absorb and oxidize 
    plutonium and uranium while leaving out other elements;
    (b) adding a reducing (de-oxidizing) agent that reduces (de-oxidizes) only the
    oxidation state of plutonium to one that is insoluble
    and thus separable; and finally (c) extracting the plutonium from the mixture. In this way
    the oxidized plutonium and uranium are separated and can be reduced back to reusable radionuclides.\cite{lb01} 
    The extracted plutonium then returns to the reactor and transforms into short-lived fission products
    by fission. Un-reprocessed fuel usually takes 10 times as much time to reach the safe
    activity level (\sim$10^5$ years) as reprocessed fuel (\sim$10^4$ years). Moreover, since fission products
    make up a small portion of the spent fuel, reprocessing greatly reduces the mass of the waste. The reprocessed
    waste can be turned into liquid form, mixed with frit (the substance from which glass is made) and then
    vitrified (made into glass). The vitrification process immobilizes the radioactive particles and makes
    the waste much more safe and manageable, but vitrification can only be applied to reprocessed spent fuel.

    It is worthwhile to note that the PUREX processed was designed specifically to separate plutonium for
    bombs. Therefore, literally any other reprocessing approach would be less prone to proliferation.
    Researchers are actively investigating alternatives to the PUREX technology. For example, the UREX+ (for
    uranium extraction) technique extracts the uranium but leaves the plutonium mixed with other
    radioactive materials. Another popular alternative, pyrochemical reprocessing, inserts giant electrodes
    into the chemical bath of nuclear waste and electroplates the plutonium and other transuranic elements onto
    the electrode. However, none of these alternatives have been put into practice due to the high construction
    and operating costs of reprocessing facilities.\cite{aa12}

@@

    \subsection{Dry cask storage}
    Spent fuel pools have very limited capacities. 
    Overground dry cask storage provides a low-cost temporary storage before the spent fuel is disposed of.
    After cooling in the spent fuel pools for at least a year, the spent fuel can be stored in dry casks filled with inert gas. 
The dry casks are often metal cylinders welded to prevent leakage. Each cask is surrounded by
    additional steel and concrete to provide shielding.
    The heat from the cask is removed by passive
    cooling (air circulation). The NRC is hoping that the casks will
    last safely for as long as 100 years, yet corrosions and cracking occur in 30 years
    or less.\cite{aa12} As such, without the prospect of a permanent geological repository,
    overground dry cask storage does not seem to be a long-term solution. The doubts were expressed
    during the Nuclear Waste Technical Review Board 2009 meeting \cite{nwtb09, aa12}
    \begin{displayquote}
    Leaving the spent fuel on-site for extended periods of time was never intended and is not
    responsible. (Dry cask manufacturer)

    It is not ethical, basically to plan for long-term storage without pursuing a well defined repository
    program. (Department of Energy employee speaking as a citizen)
    \end{displayquote}
    Even more striking is the fact that more than $75\%$ of the spent fuel are still stored in densely
    packed pools
    that are close to or over their designed limits.\cite{a11,aa12}
    Therefore, there is clearly a need for a long-term disposal plan, which we will explore in
    the next section.\\
    \red{(A lot more drama with the NRC going on here, loss of credibility.)}

\section{Disposal of HLW}
\label{sec:disposal}
    The time frame of nuclear waste disposal is about $10^4-10^6$ years, after which the radioactivity
    of the waste is predicted to decline to a safe level.
    Deep geological repositories that are stable over timescales naturally offer a good option for disposal.
    It is accepted that geological repositories in a stable environment over 300 meters below ground
    is safer than indefinite overground storage of HLW, reprocessed or not.\cite{fmr11}
    In this section, we first state the requirements for a HLW geological disposal solution, and then
    we look at  deep geological repository, and its alternatives (or variations): deep boreholes storage
    and reversible repository.
    \subsection{Requirements for HLW disposal}
    \subsection{Survey of possibilities}
    transmutation, ocean floor, deep geological disposal
    \subsection{Deep borehole disposal}
        \subsection{Deep geological repository: Yucca Mountain}
        \subsubsection{Risk}
    \subsection{Reversible disposal sites}

\section{Discussions}
meep

\begin{acknowledgments}
Thanks Joel and peeps. Arjendu and M. V. Ramana.
\end{acknowledgments}
